---
output: github_document
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# injurymatrix

The R package **injurymatrix** purpose is to facilitate the use of the [ICD-10-CM injury matrix](https://www.cdc.gov/nchs/injury/injury_tools.htm) in data analysis.  

The package provides two main functions: `matrix_intent()` and `matrix_mechanism()` to add respectively intent and mechanism of injury to the inputed data. The analyst has the option to use keywords to limit the query of intent or mechanism.
Try *`?matrix_mechanism`* and *`?matrix_intent`* for more information on those two functions. 

## Installation

To install and load the **injurymatrix** package into your working environment:

* Install the devtools package: `install.packages("devtools")`  
* Install the injurymatrix package: `devtools::install_github("epinotes/injurymatrix")`  
* Load the package: `library(injurymatrix)`  

## Examples  

```{r}
# loading relevant packages  

library(tidyverse)
library(injurymatrix)

```



```{r}
# check the content of the dataset used in the examples below.   

set.seed(11)

icd10cm_data150 %>% sample_n(10)

# get the indices of the columns with ICD-10_CM. 

grep("diag|ecode", names(icd10cm_data150), ignore.case = T)

# The indices will be used as arguments in the following functions.  

```

### Using `matrix_intent()`

* Without keyword submitted, all the five injury intents are added to the data.  
* With keywords (the partial name of the intent will suffice) only the matching intents will be added to the dataset.    


```{r}
# ?matrix_intent for more information

# No keyword is used

results_1 <- icd10cm_data150 %>% 
  matrix_intent(inj_col = c(2:6))

results_1

# table of the injury intent from result_1  

results_1 %>%
select(-diagnosis_1:-ecode2) %>%
pivot_longer(cols = -uid,
             names_to = "intent",
             values_to = "count") %>%
group_by(intent) %>%
summarise_at(vars(count), sum)

# Keywords used  

results_2 <- icd10cm_data150 %>% 
  matrix_intent(inj_col = c(2:6), "unintent", "undeterm")
  
results_2

```

### Using `matrix_mechanism()`  

* Without keyword submitted, all the 31 injury mechanisms are added to the data.  
* With keywords (the partial name of the mechanism will suffice) only the matching mechanisms will be added to the dataset.    


```{r}

# ?matrix_mechanism for more information 

# No keyword 

results_3 <- icd10cm_data150 %>% 
  matrix_mechanism(inj_col = c(2:6))
  
results_3

# Keyword used

results_4 <- icd10cm_data150 %>% 
  matrix_mechanism(inj_col = c(2:6), "firearm", "fall", "pierce")
  
results_4

# table of selected mechanisms from result_4  

results_4 %>%
select(-diagnosis_1:-ecode2) %>%
pivot_longer(cols = -uid,
             names_to = "mechanism",
             values_to = "count") %>%
group_by(mechanism) %>%
summarise_at(vars(count), sum)

```

## Data included in the `injurymatrix` package

Exploring the datasets below that provided the necessary information used by the functions described above. Run the following lines of code to get more details on the datasets. 

`library(injurymatix)`   
`?icd10cm_mech_regex` # matrix collapsed to the mechanisms    
`?icd10cm_intent_regex` # matrix collapsed to the intents   
`?injury_matrix_all` # the full matrix   

