---
output: github_document
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# injurymatrix

The R package **[injurymatrix](https://epinotes.github.io/injurymatrix/)** purpose is to facilitate the use of the **ICD-10-CM injury matrix**in data analysis. The online matrices were updated in October 2020( [ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/injury/tools/](ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/injury/tools/)).

The package provides three main functions: `matrix_intent()`, `matrix_mechanism()` and `matrix_intent_mechanism()` to add respectively intent only, mechanism only, and combination of intent and mechanism of injury to the inputed data. The analyst has the option to use keywords to limit the query of intent or mechanism.
Try *`?matrix_intent`*, *`?matrix_mechanism`* and *`?matrix_intent_mechanism`* for more information on those functions. 
There are more capabilities in the package [useicd10cm](https://github.com/epinotes/useicd10cm) to consider.  

## Installation

To install and load the **injurymatrix** package into your working environment:

* Install the devtools package: `install.packages("devtools")`  
* Install the injurymatrix package: `devtools::install_github("epinotes/injurymatrix")`  
* Load the package: `library(injurymatrix)`  

## Examples  

```{r}
# loading relevant packages  

library(tidyverse)
library(injurymatrix)

```



```{r}
# check the content of the dataset used in the examples below.   

set.seed(11)

icd10cm_data150 %>% sample_n(10)

# get the indices of the columns with ICD-10_CM. 

grep("diag|ecode", names(icd10cm_data150), ignore.case = T)

# The indices will be used as arguments in the following functions.  

```

### Using `matrix_intent()`

* Without keyword submitted, all the five injury intents are added to the data.  
* With keywords (the partial name of the intent will suffice) only the matching intents will be added to the dataset.    


```{r}
# ?matrix_intent for more information

# No keyword is used

results_1 <- icd10cm_data150 %>% 
  matrix_intent(inj_col = c(2:6))

results_1

# table of the injury intent from result_1  

results_1 %>%
select(-diagnosis_1:-ecode2) %>%
pivot_longer(cols = -uid,
             names_to = "intent",
             values_to = "count") %>%
group_by(intent) %>%
summarise_at(vars(count), sum)

# Keywords used  

results_2 <- icd10cm_data150 %>% 
  matrix_intent(inj_col = c(2:6), "unintent", "undeterm")
  
results_2

```

### Using `matrix_mechanism()`  

* Without keyword submitted, all the 33 injury mechanisms are added to the data.  
* With keywords (the partial name of the mechanism will suffice) only the matching mechanisms will be added to the dataset.    


```{r}

# ?matrix_mechanism for more information 

# No keyword 

results_3 <- icd10cm_data150 %>% 
  matrix_mechanism(inj_col = c(2:6))
  
results_3

# Keyword used

results_4 <- icd10cm_data150 %>% 
  matrix_mechanism(inj_col = c(2:6), "drug", "fall", "pierce")
  
results_4

# table of selected mechanisms from result_4  

results_4 %>%
select(-diagnosis_1:-ecode2) %>%
pivot_longer(cols = -uid,
             names_to = "mechanism",
             values_to = "count") %>%
group_by(mechanism) %>%
summarise_at(vars(count), sum)

```  


### Using `matrix_intent_mechanism()`  

* Without keyword submitted, all the 92 injury intents and mechanisms combined are added to the data.  
* With keywords (the partial name of the mechanism or intent will suffice) only the matching combination of intent and mechanisms will be added to the dataset.    


```{r}

# ?matrix_mechanism for more information 

# No keyword 

results_5 <- icd10cm_data150 %>% 
  matrix_intent_mechanism(inj_col = c(2:6))
  
results_5

# Keyword used

results_6 <- icd10cm_data150 %>% 
  matrix_intent_mechanism(inj_col = c(2:6), "Poisoning_Drug")
  
results_6

# table of selected mechanisms from result_4  

results_6 %>%
select(-diagnosis_1:-ecode2) %>%
pivot_longer(cols = -uid,
             names_to = "intent_mechanism",
             values_to = "count") %>%
group_by(intent_mechanism) %>%
summarise_at(vars(count), sum)

```  


### Create A column of first valid external cause 

This example illustrates how to create a first valid external cause field.  

  
```{r}
icd10cm__external_cause_ <- "(^[VWX]\\d....|(?!(Y0[79]))Y[0-3]....|Y07.{1,3}|Y09|(T3[679]9|T414|T427|T4[3579]9)[1-4].|(?!(T3[679]9|T414|T427|T4[3579]9))(T3[6-9]|T4[0-9]|T50)..[1-4]|T1491.{0,1}|(T1[5-9]|T5[1-9]|T6[0-5]|T7[1346])...|T75[0-3]..)(A|$)"
```  

The function `icd_first_valid_regex()` in combination with the regular expression above,  `icd10cm__external_cause_` (It is in the [CSTE Toolkit](https://resources.cste.org/ICD-10-CM/Standardized%20Validation%20Datasets/Other%20Useful%20ICD-10-CM%20Regular%20Expressions.pdf)) will create the first valid external cause field.


```{r}
results_7 <- icd10cm_data150 %>% 
  mutate(ex_cause1 = icd_first_valid_regex(., colvec = c(2:6), 
                                           pattern = icd10cm__external_cause_))

results_7

```

Adding selected intents to *results_7* by using the new *ex_cause1* only:  


```{r}
results_7 %>% 
  matrix_intent(inj_col = "ex_cause1", 
                "unintent", "undeterm")
```


### Filter A Dataset With Valid External Cause Of Injury ICD 10 CM    


Using the function `matrix_valid_external()` providing the columns with the icd-10-cm of interest.  

```{r}
# create a new binary variable "valid_external"  

results_8 <- icd10cm_data150 %>% 
   matrix_valid_external(c(2:6))

results_8

# count the number of records with valid external cause of injury
results_8 %>% 
  count(valid_external)

# Subset the dataset to the 136 records with valid external causes of injury.

results_8 %>% 
  filter(valid_external == 1)

```


## Data included in the `injurymatrix` package

Exploring the datasets below that provided the necessary information used by the functions described above. Run the following lines of code to get more details on the datasets. 

`library(injurymatix)`   
`?icd10cm_mech_regex` # matrix collapsed to the 33 mechanisms    
`?icd10cm_intent_regex` # matrix collapsed to the 5 intents 
`?icd10cm_intent_mech_regex` # matrix of the 92 combinations of intent and mechanism
`?injury_matrix_all` # the full matrix of 3,655 entries ICD-10-CM of external causes of injury

